In this chapter we will give a background of the relevant economic theory for this thesis. We aim to formally introduce a central theory in economics which is General Equilibrium Theory. Along with it, we hope to expose the reader with Physics background to the usual concerns of Economics. We also discuss the common criticisms about General Equilibrium Theory, along with more recent work to solve some of the problems discussed.

General Equilibrium Theory is a mature field of Economics \cite{Arrow54, mascolell, mckenzie} which aims to
characterize the existence and properties of equilibria in certain market settings. Economic systems are frequently assumed to have
actors with opposing goals: the owner of a good wants to sell it for the highest possible price, while its potential buyers would like to
purchase it for as low as possible. Fishermen would like to catch as many fishes as possible and expect their peers to not also overdo
it otherwise they may extinguish the oceans. Businessmen would like to price their products in such a way that as many people as possible buy it for as high as possible. A couple in vacation would like to buy a plane ticket to Paris as cheap as possible as long as the layover is not longer than a couple of hours. One expects that economies and markets composed of many such actors trying to get their way will converge to a certain steady state. The main goal of General Equilibrium Theory is trying to characterize these equilibrium states in a rigorous manner. In this sense, it's a microeconomic theory, because it explains macrobehavior from the incentives (or micromotives \cite{SchellingBook}) of agents.

While the work in this thesis doesn't strictly aim at studying economies as they are described in General Equilibrium Theory, its
purposes and questions are very similar to the ones physicists usually ask when studying a complex system: namely, the characterization
of its equilibrium state. It's important, therefore, to have a deeper understanding of how this is done in economics, what are its main
concerns and assumptions, if only to realize the two field have very similar goals.

\section{A Brief Exposition}

The exposition in this chapter is mostly adapted and simplified from Mas Colell's seminal book on Microeconomic Theory \cite{mascolell} and will be considerably more formal than the rest of this thesis due to the way the discipline is commonly studied.

In General Equilibrium Theory, an economy is defined through the following components: we assume there are $J$ consumers, $N$ firms and $M$
goods in a market. Each consumer has a consumption set $X_j$ which contains all possible consumption bundles $x_j = (x_j^1, \ldots x_j^M)$ that the
consumer has access to, i.e., each bundle $x_j$ is a $M-$dimensional vector with nonnegative entries (we are assuming he cannot consume a
negative amount of a good). $X_j$ is limited by ``physical'' constraints, such as no access to water, iron or bread, but not monetary
constraints, which we will deal with later.

The consumer also has an utility function $U_j(x)$ that takes every element $x_j \in X_j$ to a real number, representing how much the
consumer values each bundle of his consumption set. This allows us to define define a preference relationship over the elements in $X_j$
(i.e., if the consumer prefers bundle $x$ to $x'$), which is complete\footnote{For every $x, x' \in X_j$, either
  $U_j(x) \geq U_j(x')$ or $U_j(x) \leq U_j(x')$.} and transitive\footnote{For every $x, y, z \in X_j$, if
  $U_j(x) \geq U_j(y)$ and $U_j(y) \geq U_j(z)$, then $U_j(x) \geq U_j(z)$.}, two standard requirements in Economics for
rational behavior.

Finally, the consumer is also endowed with an initial bundle of goods $\omega_j = (\omega_j^1, \ldots, \omega_j^M)$, $\omega_j^\mu \geq 0$
which will define his budget given a set of prices for the goods and will constraint his choices on $X_j$. His initial budget is defined in terms of a bundle of goods for two reasons: first, we want the economy as a whole to be unaffected by the scales of prices, so if all prices are multiplied by a constant factor the consumer's purchasing power remains the same. Second, the theory described \textit{general} equilibriums, which are assumed to be valid for closed markets. All goods used as input have to come from somewhere inside the economy, and the consumer acts as the provider of raw material.

Each firm $i$ has a production set $\Xi_i$ of technologies $\xi_i = (\xi_i^1, \ldots, \xi_i^M)$ that it is able to operate. Unlike
consumption bundles, which are final allocations and therefore must be nonnegative, technologies can be any real number: the negative entries
are inputs and the positive entries are outputs in the technology's operation. $\Xi_i$ is also limited only by ``physical'' constraints, not
by monetary constraints. A firm that has $\xi_i = (-1, 2)$ in its production set is able to transform one unit of good 1 into two units
of good 2. It won't necessarily be able to transform two units of good 1 into four units of good 2, for that it must also have $\xi'_i = (-2,
4)$ in $\Xi_i$. It might be the case, for example, that companies get more efficient with production and therefore it might have $\xi''_i = (-2, 6)$ in its production set.

In General Equilibrium Theory, an economy is formally defined as the tuple

\begin{equation}
  E = \left(\{(X_j,U_j)\}_{j=1}^J,
    \{\Xi_i\}_{i=1}^N, \{\omega_j\}_{j=1}^J \right).\label{eq:econGE}
\end{equation}

One of the theory's assumptions is that the economy described is \textbf{complete}, which means every agent can exchange every good with no transaction costs and complete information about the firm's technologies, other consumer's consumption, etc. Also, a good $\mu$ contains all the possible information that a consumer would take into account when making his choice. That is, among the space of goods we could have ``umbrella'' and ``chocolate'', or we could also have ``an umbrella on August 13th, 2016 in São Paulo with 50\% chance of rain'' and ``an umbrella on December 12th, 2016 in Chicago with 90\% chance of rain''. There is no hidden information, so the consumer is never in doubt of whether the good or company is actually reliable or any such considerations.

It's also assumed that agents are \textbf{price-takers}, that is, they are unable to affect the market prices and therefore have to take them as a given. The prices of the goods are given by a $M-$dimensional vector $p = (p_1, \ldots, p_M)$, where each price is a strictly positive quantity, i.e., $p_\mu>0$ for all $\mu$. This carries the restriction that goods have global prices, which is consistent with the completeness assumption: there is no reason why the market prices should be different for certain consumers or firms if they have complete knowledge and no transaction costs.

With a price vector $p$ defined, we say the consumer $j$ has a budget $B_j = p\cdot \omega_j$, which is the monetary value of his initial endowment. Any bundle he chooses to purchase will cost him $p\cdot x_j$. His objective, therefore, is to find the best bundle $x_j$ he is able to afford, that is:

\begin{equation}
  \label{eq:consumer_obj}
  \max_{x_j \in X_j} U(x_j), \quad \text{s.t. } p\cdot x_j \leq p\cdot \omega_j
\end{equation}

The firms, on the other hand, have an operating profit for each technology given by $p \cdot \xi_i$, which is how much money they earn by selling their outputs ($\xi_i^\mu > 0$) minus how much they spend purchasing the inputs ($\xi_i^\mu < 0$). Their objective is to maximize their profits, that is:

\begin{equation}
  \label{eq:firm_obj}
  \max_{\xi_i \in \Xi_i} p \cdot \xi_i
\end{equation}

With these ingredients laid out, we define an \textbf{allocation} of the economy as a set of specific choices for consumption bundles and technologies, i.e., an allocation $a$ of an economy $E$ is

\begin{equation}
  \label{eq:GET_3}
  a = (x_1, \ldots, x_J, \xi_1, \ldots, \xi_N), \, x_j \in X_j, \,
  \xi_i \in \Xi_i
\end{equation}

We are considering closed economies and therefore all that is produced must come from the initial endowments and end up part of the consumers' final bundles. We therefore say an allocation is \textbf{feasible} if it satisfies \textbf{market clearing} for all the goods:

\begin{equation}
  \label{eq:GET_market_clearing}
  \sum_{j=1}^J x^\mu_j = \sum_{j=1}^J \omega_j^\mu + \sum_{i=1}^N
  \xi_i^\mu, \quad \forall \, \mu \in \{1,\ldots, M\}
\end{equation}

This is a strong condition which couples many quantities in the economy. In particular, if we multiply both sides of the equation by $p^\mu$ and sum over $\mu$ we get, in vector notation,

\begin{equation}
  \label{eq:GET_price_market_clearing}
  \sum_{j = 1}^J p \cdot (x_j - \omega_j) = \sum_{i=1}^N p\cdot \xi_i
\end{equation}

The left-handside is the leftover money the consumers have after making their choice of consumption, also called the value of excess demand, whereas the right-handside is the firms aggregate profit, also known as the value of excess supply. Because we assume that the consumer may not spend more than his budget, the value of each consumer's individual excess demand has to be non positive. Simultaneously, if we assume that firms always have $\xi_i = 0$ in their production set, i.e., we assume that they can always opt to not produce at all and leave the market, then the value of excess supply for each firm has to be non negative. Because they must be equal, we conclude that in an economy for which market clearing holds, the consumer spends all his available budget and firms all get zero profit, a result known as \textbf{Walras' Law}. 

Given a set of feasible allocations $\{a_k\}$, we may wonder if there is any allocation we desire most over the others. This of course depends on the criteria we use to judge them: we may like allocations with less inequality, with the most aggregate utility, with the smallest minimum utility, etc. Economists opt to use one particular condition which is called \textbf{Pareto optmality}.

Intuitively, a \textbf{Pareto optimal} (or \textbf{Pareto efficient}) allocation is one in which you can't make a consumer better without making another consumer worse off. The idea is that, a non Pareto optimal allocation has some waste in it: one could change the consumption bundles in order to increase some utilities and no other consumer would complain. Because firms have zero profit in feasible allocations, they wouldn't mind the change. 

More formally, a feasible allocation $a = (x, \xi)$ is said to be \textbf{Pareto optimal} if there is no other allocation that \textbf{Pareto dominates} it, that is, no allocation $a' = (x', \xi')$ such that $U(x_j') \geq U(x_j)$ for all $j$ and $U(x_j') > U(x_j)$ for at least one $j$.

The Pareto optimality concept therefore defines a socially desirable outcome in a ``non-controversial'' way, by definition no agent in the economy would have a problem with policies or actions taken to make it more Pareto efficient. However, it says nothing about equality: an allocation in which one consumer has all the goods and no other consumer has any goods is Pareto optimal.

We finally arrive at the concept of equilibrium in an economy. A \textbf{Walrasian equilibrium} (or competitive equilibrium or simply equilibrium) in an economy E is an allocation $(x^\ast, \xi^\ast)$ and a price vector $p$ such that

\begin{enumerate}
\item Every firm $i$ maximizes its profits in its production set
  $\Xi_i$, that is
  \begin{equation}
    \label{eq:GET_4}
    p\cdot \xi_i^\ast \geq p\cdot \xi_i, \quad \forall \xi_i \in
    \Xi_i,\quad \forall i \in \{1, \ldots, N\} 
  \end{equation}
\item Every consumer $j$ maximizes his utility in his consumption set
  $X_j$, that is
  \begin{equation}
    \label{eq:GET_1}
    U(x_j^\ast) \geq U(x_j), \quad \forall x_j \in X_j, \quad \forall
    j \in \{1,\ldots, J\}
  \end{equation}
\item The allocation $(x^\ast, \xi^\ast)$ is feasible, that is,
  \begin{equation}
    \label{eq:GET_2}
    \sum_{i=j}^J x_j^\ast = \sum_{j=1}^J \omega_j + \sum_{i=1}^N \xi_i^\ast
  \end{equation}
\end{enumerate}

The Walrasian equilibrium is essentially a pair allocation - prices such that all optimization problems are solved at once. Although we have not mentioned any dynamics in this economy, it's considered an equilibrium because all agents are as satisfied as possible with their allocation given the prices, which we have assumed to be global and unchangeable by any agent's action. This is not exactly a definition of equilibrium as used in Physics, but we will discuss this point later. For now, we point out that a Walrasian equilibrium is in some sense stable.

We have thus defined two desirable properties of an allocation: efficiency and equilibrium. The fundamental results of General Equilibrium Theory are the \textbf{Welfare Theorems}, which define the conditions for which an equilibrium is Pareto optimal and for when a specific Pareto optimal allocation is an Walrasian equilibrium.

The \textbf{First Fundamental Welfare Theorem} asserts that if the consumers have a continuous utility function on $X_j$\footnote{The actual theorem   asserts a weaker condition, that the preferences be locally nonsatiated, that is, for every $x \in X$, there is an $x' \in X$   such that $\|x - x'\| < \varepsilon$ and $x'$ is preferred to $x$.}, then all Walrasian equilibria are Pareto optimal. This result is simple yet useful, because it tells us that if our economy is in equilibrium, we don't have to care about checking if it's efficient. The violation is also important: if a given economy we are studying is in an inefficient equilibrium, then it must be that one of the theorem's condition was violated. This sheds light in where to look for market failures. We remind the reader, however, that some extra strong assumptions were made for the economies described by this theorem, namely, completeness of market and global prices that no single agent is capable of influencing.

The \textbf{Second Fundamental Welfare Theorem} requires extra assumptions: it affirms that if an economy satisfies the conditions of the first fundamental theorem, the utility functions $U_j$ and all sets $X_j$, $Y_i$ are convex and if we are able to redistribute the initial endowments at will, while keeping the total amount $\sum_{j=1}^J \omega_j$ constant, then for every Pareto efficient allocation there exists a wealth allocation $\omega$ and price vector $p^\ast$ such that $(x^\ast, \xi^\ast, p^\ast)$ is a Walrasian equilibrium.

The second theorem is considerably more interesting than the first one: any Pareto optimal allocation we would like in an economy can be an equilibrium given the appropriate price vector and a possible wealth transfer, albeit under a stronger set of conditions. It serves both as a benchmark, because we know what we can expect from markets at their "optimal conditions", but also as a warning: we can only guarantee that an efficient allocation will be an equilibrium under a very strong set of requirements.


\section{Limitations of General Equilibrium}

A conspicuous element was missing from the exposition above: there are no rules for the dynamics of the economies described above. The prices are taken as a given, as are the consumer and firm choices. What happens if a firm closes? What happens if a new firm appears? The equilibrium is simply ``recalculated'' and the economy moves to the new one?

Indeed, this is a long standing criticism to General Equilibrium Theory. Walras proposed it as a process of \textbf{tatônnement}\footnote{From ``trial and error'' in french}: a central figure, known as the Walrasian auctioneer, suggests a price and asks all the firms and consumers how much would they like to produce and buy at these given prices, but without any transaction taking place at out of equilibrium prices. The auctioneer updates the prices in the direction of diminishing excess demand or supply, a gradient descent process, until equilibrium is reached, at which point transactions finally take place. The auctioneer must also have a way of guaranteeing that agents will be price takers: it must either be able to monitor and enforce all transactions or buy and sell arbitrary amounts of every good at their equilibrium prices.

It is clear that such process is very convoluted. Chiefly, this auctioneer figure doesn't exist in most decentralized markets: goods are traded at agreed prices by both parts, which do not wait until their transaction is authorized by some central authority. Even if there were auctioneers, such authority would require an infinitely large computational capability to compute the excess demand and supply of every consumer and firm and for every good in a modern economy \cite{axtell05, Roughgarden10}. Worst of all, even if there was such central figure with such an arbitrary large amount of computing power, not all price updating dynamics are guaranteed to converge \cite{Gintis07}. Finally, even if it converges, we have no assurance that it will converge in finite time.

These are all well known and acknowledged shortcomings of the theory. Some areas of Economics, such as the Schumpeterian Economics \cite{Schumpeter39}, eschew the ideal of an static equilibrium altogether, studying the pattern of changes from certain evolutionary rules instead. These ideas are also popular amongst physicists \cite{thurner}.

However, the usefulness of General Equilibrium Theory in Economics stands not from practical applications, but as a benchmarking tool for real world policies. Their conclusions are mathematically precise and correct. So when faced with an economic equilibrium that is not Pareto efficient, then by definition it must be because one of the Second Fundamental Welfare Theorem conditions was violated, and therefore one knows where to look for inefficiencies to try and fix it.

Decades after the introduction of the Welfare Theorems, the field of Applied General Equilibrium arose as a way to compute equilibria and explore them for policy decisions \cite{shoven1984, shoven1992}. It used real world data to calibrate production and utility functions along with iterative schemes to calculate the equilibrium prices \cite{scarf1967}. From there, it could be used to explore the impact of policy changes in more complex settings.


More recently, these limitations have been tackled by the Dynamic Stochastic General Equilibrium (DSGE) models, that have origin in the work by Kydland and Prescott \cite{kydland1982}. In that seminal paper, the authors calculate the evolution in time of macroeconomic variables such as economic output over time by treating it as the trajectory in time of a microeconomic general equilibrium problem with stochastic elements. This has led to a new class of economic models based on this approach of using dynamical systems with stochastic terms and inferring the parameters from real world data \cite{christiano2005, smets2007}, which are employed today by institutions such as the European Central Bank for policy analysis \cite{smets2003}.

However, what all of theses models have in common is that they all look for equilibria in the Classical Mechanics sense of Physics. Generally speaking, an equilibrium state in Economics is when all incentives cancel each other out and no actor has the desire or the possibility of deviating from the current configuration, which is an approach that heavily draws from Classical Mechanics in Physics and dynamical systems in Mathematics. In fact, Stephen Smale even wrote a paper on the dynamics of General Equilibrium \cite{smale1976} trying to tackle some of the open dynamical questions discussed earlier, employing modern mathematical theory of dynamical systems.

This Classical Mechanics approach has limitations: due to the need of having one exact solution for the equilibrium configuration, economic models usually employ some standard simplifications to make problems tractable. The most common of which is the representative agent, in which a single agent represents all consumers, another represents all firms, all the government policies, etc, and his objective functions are considered as the average of an heterogeneous population, with the different parts of the economy interacting through these "average demands".

Trimming down a problem to its bare essential features isn't exactly unknown in Physics. However, one must be careful not to change the problem completely when doing so. In Statistical Physics, it is widely known that it is precisely the interaction between a large number of particles that generate rich phenomena such as phase transitions, in which the average quantities of a system go through a discontinuous transition. From that point of view, it seems a waste of opportunity to treat the demand function for all consumers as a continuous and well behaved function. This has been pointed out both by economists \cite{Kirman92, KirmanBook} and physicists \cite{Bouchaud13, Bouchaud07} and probably the most common criticism of Economic Theory is that it has done a poor job in predicting major crisis as recently as the 2007 - 2008 burst of the housing bubble in the US, which generated a new wave of criticism for the lack of foresight \cite{BouchaudCrisis}. Using interacting agents allows for models in which crisis and transitions come from endogenous rules of interaction instead of arbitrary adhoc modelling.

But perhaps the biggest difference in the approaches is that, in Statistical Mechanics, an equilibrium is defined as an ensemble of possible configurations that appear with certain probabilities, and instead of trying to find the exact numbers that equilibrate the system, one looks for the average of macroscopic quantities. This allows for the interpretation of fluctuations as a natural phenomena, due to the nature of the uncertainty involved in inferring a complex system, as we will describe in the next chapter. Then, one can truly define the behavior of microscopic interactions, as opposed to representative agents, and say something about the aggregate.

Finally, because fluctuations around a minimum are a fundamental part of the Statistical Mechanics framework, they allow us to have a principled way of modeling agents that don't necessarily act in an optimal way. Ever since the work of Tversky and Kahneman \cite{Tversky74, Kahneman79, Tversky81}, economic mainstream has paid more and more attention to the empirical evidence that real life decisions are not always optimal. While this has been known for a long time, traditional Economic Theory has always treated optimality in choice as a good approximation to real life scenarios, akin to disregarding air drag in a Classical Mechanics problem. However, the amount of evidence that human beings consistently make suboptimal choices has increased the need for alternative approaches. In Economic Theory, suboptimal behavior is usually introduced as a stochastic shock in the consumer optimization problem or through other adhoc heuristics. A Statistical Mechanics approach can contribute to this topic by offering a natural way of dealing with these fluctuations, as we will show in Chapters \ref{cha:stat_mech} and \ref{cha:inefficient}. Later, in Chapter \ref{cha:inequality} we will show that we can get valuable insights into statistical properties of the economy if we assume no rationality at all -- an agent that chooses at random.