% book example for classicthesis.sty
\PassOptionsToPackage{hyphens}{url}\documentclass[
  % Replace twoside with oneside if you are printing your thesis on a single side
  % of the paper, or for viewing on screen.
  %oneside,
  twoside,
  11pt, a4paper,
  footinclude=true,
  headinclude=true,
  cleardoublepage=empty
]{scrbook}
% \usepackage[T1]{fontenc}
\usepackage[linedheaders,parts,pdfspacing]{classicthesis}
% \usepackage{lmodern}
\usepackage{acronym}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[annote,fixlanguage]{babelbib}
\usepackage[footnotesize]{caption}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{dsfont}
\usepackage{mdwlist}
\usepackage{mathrsfs}
\usepackage{nicefrac}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{subfig}
\usepackage[left=1.25in,right=1.25in]{geometry}
\DeclareMathOperator{\sen}{sen}  \newcommand{\del}{\partial}
\newcommand{\R}{\mathds{R}} \newcommand{\C}{\mathds{C}}
\newcommand{\N}{\mathds{N}} \newcommand{\Q}{\mathds{Q}}
\newcommand{\Z}{\mathds{Z}}
\DeclareMathOperator*{\argmin}{arg\,min}\DeclareMathOperator*{\argmax}{arg\,max} \DeclareMathOperator\erfc{erfc}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}


\begin{document}
\begin{titlepage}
  \begin{center}

    \textsc{\Large Universidade de São Paulo} \\[0.1cm]
    \textsc{\Large Instituto de Física} \\ [4cm]
    
{\Large \textbf{Mecânica Estatística de Sistemas Econômicos}} \\ [1cm]

{\Large João Pedro Jericó} \\[3cm]
  \end{center}


  \begin{flushleft}
  \hspace{9cm} Orientador: Renato Vicente \\ [1cm] 
\hspace{8cm}    Tese de Doutorado apresentada ao \\ \hspace{8cm}  Instituto de
Física para a
    obtenção do \\ \hspace{8cm}  título de Doutor em Ciências
  \end{flushleft}
\vspace{1cm}

\begin{flushleft}
Banca examinadora:

Prof. Dr. Renato Vicente (IME-USP) \\
Prof. Dr. Nestor Felipe Caticha Alfonso (IF-USP) \\
Prof. Dr. Fernando Tadeu Caldeira Brandt (IF-USP) \\ 
Prof. Dr. André Cavalcanti Rocha Martins (EACH-USP) \\
Prof. Dr. Masayuki Oka Hase (EACH-USP) \\
\end{flushleft}

\vfill

\begin{center}
  {\large São Paulo} \\
{\large 2016}
\end{center}
\end{titlepage}


\include{FrontBackMatter/acknowledgements}
\include{FrontBackMatter/portuguese_abstract}
\include{FrontBackMatter/english_abstract}
% \include{FrontBackMatter/dedication}
\include{FrontBackMatter/contents}



\chapter{Introduction}

\input{Introduction}



\chapter{Economics and the General Equilibrium Theory}
\label{cha:GET}
\input{GeneralEquilibriumTheory}

\chapter{Statistical Mechanics and Inference}
\label{cha:stat_mech}
\input{StatisticalMechanics}

\chapter{The Random Linear Economy model}
\label{cha:RLE}

\input{RandomLinearEconomy}


\chapter{Inefficient Consumer in a General Equilibrium Setting}
\label{cha:inefficient}

\input{Inefficient}

\chapter{Input-Output of Random Economies and Real World Data}
\label{cha:IO}

\input{InputOutput}

\chapter{When Does Inequality Freeze an Economy?}
\label{cha:inequality}
\input{Inequality}

\chapter{Conclusion}
\input{Conclusion}

\appendix

\chapter{Calculation of the Partition Function for the Random Linear
  Economy model} 
\label{app:rep}

\input{AppendixReplicaCalc}

\chapter{Definitions}
\label{app:stat}

\input{AppendixStatisticalMeasures}

\bibliography{bib}{}
\bibliographystyle{plain}
    
\end{document}