%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgements}{acknowledgements}
\chapter*{Acknowledgements}


I offer my deepest thanks to my advisor, professor Renato Vicente, for his dedication and patience throughout this PhD, for the great deal of advices and for playing a large role in my personal and professional development.

I am also very thankful to Nestor Caticha for introducing me to Statistical Mechanics with his usual excitement ten years ago and showing a motivating interest to any problems or questions I presented him with.

I thank Matteo Marsili for receiving me for an internship at the ICTP in Trieste and for the fruitful collaboration we had. Also, thanks to the friends I made in Italy: Valerio, Ryan, Daniele, Matteo, Asja, Federico, Antonio, Erica and others without whom living in another continent would have been a much daunting task. Also Bruno being there helped a lot with this too, so thanks! I am also grateful to the Abdus Salam ICTP for the hospitality in hosting me.

I specially thank André and Paulo for sharing with me the highs and lows of doing a PhD during these four years. It was tremendously important and it would have been much harder without them. The same goes for Bruno, Gabriel, Lucas, Henrique and Petre, whom I have known for ten years now and whose friendship I treasure highly, even when some of them call me at 5am fully aware that there is a 4 hour time difference between us.

I thank the people at the Statistical Physics group in IFUSP: André, Diogo, Carol, Cinthia, Jonatas, Rafael and others, for the interesting discussions, weekly seminars and friendly conversations. I also thank the people at work: Caio, Danilo, Darlan, Djalma, Luciano, Luiz, Marcelo and Paulo for their support during this last year during which I wrote this thesis. I specially thank them for the daily nag of when this thing would be done already.

I am grateful to CNPq and FAPESP for the financial support of my PhD, the latter under grant n. 2012/19521-8 and also under grant n. 2014/16045-6 for the internship abroad via the BEPE program. I also am very grateful to Dropbox for keeping all my PhD work nice and tidy in their servers. Seriously though, super helpful.

I want to thank both my families, Jericós, Andrades, Vasarinis, Quinis and Lopes, for giving me the sense of belonging and security nothing else can. Special thanks to my parents, Márcia and João Pedro, whom I will be forever indebted to for their unwavering support over all these years, giving me a safety net that allowed me to pursue my education with no restraints.

Finally, I would like most of all to thank Marina, for the life we are building together.